---
title: "Data Visualization Lab - ANSWER KEY"
author: "Answer Key"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

# Welcome to Data Visualization! 📊

This document contains complete answers for all exercises in the **Visualization Lab**.

---

# Getting Started

```{r packages}
# Load packages
library(tidyverse)  # This includes ggplot2!
library(patchwork)  # For combining charts

# Check if it loaded
cat("✓ Packages loaded successfully!\n")
```

---

# Create Sample Data

Let's make data about a lemonade stand!

```{r create-data}
# Weekly sales data
lemonade <- tibble(
  day = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"),
  cups_sold = c(25, 30, 35, 40, 50, 75, 80),
  temperature = c(70, 75, 80, 82, 85, 90, 88),
  price = c(2.00, 2.00, 2.50, 2.50, 2.50, 3.00, 3.00)
)

# Look at the data
lemonade
```

---

# PART 1: Bar Charts

## Your First Chart! 🎉

```{r first-chart}
# Basic bar chart
ggplot(data = lemonade, aes(x = day, y = cups_sold)) +
  geom_col()
```

---

## Make It Better!

```{r better-bar}
ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "skyblue") +  # Color the bars
  labs(
    title = "Lemonade Sales by Day",
    x = "Day of Week",
    y = "Cups Sold"
  )
```

**🎯 YOUR TURN:** Change the color to "coral" or "lightgreen"

```{r practice-1-1}
# ANSWER - Coral:
ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "coral") +
  labs(
    title = "Lemonade Sales by Day",
    x = "Day of Week",
    y = "Cups Sold"
  )

# ANSWER - Light Green:
ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "lightgreen") +
  labs(
    title = "Lemonade Sales by Day",
    x = "Day of Week",
    y = "Cups Sold"
  )
```

---

## Sort the Bars

```{r sorted-bar}
ggplot(lemonade, aes(x = reorder(day, cups_sold), y = cups_sold)) +
  geom_col(fill = "coral") +
  labs(title = "Sales by Day (Sorted)", x = "Day", y = "Cups Sold") +
  coord_flip()  # Make horizontal
```

**🎯 YOUR TURN:** Make a bar chart of price by day

```{r practice-1-2}
# ANSWER:
ggplot(lemonade, aes(x = day, y = price)) +
  geom_col(fill = "gold") +
  labs(
    title = "Lemonade Price by Day",
    x = "Day",
    y = "Price ($)"
  ) +
  theme_minimal()

# Alternative with sorted bars:
ggplot(lemonade, aes(x = reorder(day, price), y = price)) +
  geom_col(fill = "darkgreen") +
  labs(
    title = "Lemonade Price by Day (Sorted)",
    x = "Day",
    y = "Price ($)"
  ) +
  coord_flip()
```

---

# PART 2: Line Charts

## Basic Line Chart

```{r line-basic}
ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_line(group = 1)  # group = 1 connects all points
```

---

## Make It Better!

```{r line-better}
ggplot(lemonade, aes(x = day, y = cups_sold, group = 1)) +
  geom_line(color = "blue", size = 1.5) +  # Thicker blue line
  geom_point(color = "red", size = 3) +    # Add red dots
  labs(
    title = "Lemonade Sales Trend",
    x = "Day",
    y = "Cups Sold"
  ) +
  theme_minimal()  # Clean look
```

**🎯 YOUR TURN:** Create a line chart showing temperature over the week

```{r practice-2-1}
# ANSWER:
ggplot(lemonade, aes(x = day, y = temperature, group = 1)) +
  geom_line(color = "red", size = 1.5) +
  geom_point(color = "orange", size = 3) +
  labs(
    title = "Temperature Over the Week",
    x = "Day",
    y = "Temperature (°F)"
  ) +
  theme_minimal()

# Alternative with different styling:
ggplot(lemonade, aes(x = day, y = temperature, group = 1)) +
  geom_line(color = "darkred", size = 2) +
  geom_point(color = "yellow", size = 4) +
  labs(
    title = "Weekly Temperature Trend",
    subtitle = "Temperatures ranged from 70°F to 90°F",
    x = "Day of Week",
    y = "Temperature (°F)"
  ) +
  theme_classic()
```

---

# PART 3: Scatter Plots

## Basic Scatter Plot

```{r scatter-basic}
ggplot(lemonade, aes(x = temperature, y = cups_sold)) +
  geom_point()
```

---

## Add a Trend Line

```{r scatter-trend}
ggplot(lemonade, aes(x = temperature, y = cups_sold)) +
  geom_point(size = 4, color = "darkblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add trend line
  labs(
    title = "Sales vs Temperature",
    x = "Temperature (°F)",
    y = "Cups Sold"
  )
```

**🎯 YOUR TURN:** Make a scatter plot of price vs cups_sold

```{r practice-3-1}
# ANSWER:
ggplot(lemonade, aes(x = price, y = cups_sold)) +
  geom_point(size = 5, color = "purple") +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  labs(
    title = "Sales vs Price",
    x = "Price ($)",
    y = "Cups Sold"
  ) +
  theme_minimal()

# Alternative with different styling:
ggplot(lemonade, aes(x = price, y = cups_sold)) +
  geom_point(size = 5, color = "navy", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "pink") +
  labs(
    title = "Relationship Between Price and Sales",
    subtitle = "Higher prices are associated with higher sales (weekend effect)",
    x = "Price per Cup ($)",
    y = "Cups Sold"
  ) +
  theme_light()
```

---

# PART 4: Colors and Themes

## Adding Colors by Category

```{r color-by-category}
# Add a weekend column
lemonade <- lemonade %>%
  mutate(
    is_weekend = if_else(day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
  )

# Color by weekend
ggplot(lemonade, aes(x = day, y = cups_sold, fill = is_weekend)) +
  geom_col() +
  labs(title = "Sales: Weekday vs Weekend", fill = "Day Type")
```

---

## Built-in Themes

```{r themes}
# Try different themes!
p <- ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "steelblue") +
  labs(title = "Lemonade Sales")

# Theme 1: Minimal (clean)
p + theme_minimal()

# Theme 2: Classic (traditional)
p + theme_classic()

# Theme 3: Dark
p + theme_dark()
```

**🎯 YOUR TURN:** Try `theme_bw()` or `theme_light()`

```{r practice-4-1}
# ANSWER - theme_bw():
ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "steelblue") +
  labs(title = "Lemonade Sales") +
  theme_bw()

# ANSWER - theme_light():
ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "steelblue") +
  labs(title = "Lemonade Sales") +
  theme_light()

# Bonus - combining both concepts:
ggplot(lemonade, aes(x = day, y = cups_sold, fill = is_weekend)) +
  geom_col() +
  labs(
    title = "Lemonade Sales by Day Type",
    x = "Day",
    y = "Cups Sold",
    fill = "Day Type"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("Weekday" = "lightblue", "Weekend" = "coral"))
```

---

# PART 5: Multiple Charts Together

## Create Multiple Simple Charts

```{r multiple-charts}
# Chart 1: Bar chart
p1 <- ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "coral") +
  labs(title = "Cups Sold") +
  theme_minimal()

# Chart 2: Line chart
p2 <- ggplot(lemonade, aes(x = day, y = temperature, group = 1)) +
  geom_line(color = "red", size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Temperature") +
  theme_minimal()

# Put them side by side
p1 | p2
```

---

## Stack Charts Vertically

```{r stack-charts}
# Put one above the other
p1 / p2
```

**🎯 YOUR TURN:** Create any two charts and combine them

```{r practice-5-1}
# ANSWER:

# Chart 1: Price line chart
price_chart <- ggplot(lemonade, aes(x = day, y = price, group = 1)) +
  geom_line(color = "darkgreen", size = 1.5) +
  geom_point(color = "gold", size = 4) +
  labs(title = "Price Trend", y = "Price ($)") +
  theme_minimal()

# Chart 2: Sales vs Temperature scatter
temp_chart <- ggplot(lemonade, aes(x = temperature, y = cups_sold)) +
  geom_point(size = 5, color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Sales vs Temperature", x = "Temp (°F)", y = "Cups Sold") +
  theme_minimal()

# Side by side
price_chart | temp_chart

# Stacked
# price_chart / temp_chart

# Bonus - more complex layout:
p1 <- ggplot(lemonade, aes(x = day, y = cups_sold, fill = is_weekend)) +
  geom_col() +
  labs(title = "Sales by Day Type") +
  theme_minimal()

p2 <- ggplot(lemonade, aes(x = day, y = temperature, group = 1)) +
  geom_line(color = "red", size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Temperature") +
  theme_minimal()

p3 <- ggplot(lemonade, aes(x = temperature, y = cups_sold)) +
  geom_point(size = 4, color = "purple") +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Temp vs Sales") +
  theme_minimal()

# Complex layout
(p1 | p2) / p3
```

---

# PART 6: Saving Your Charts

## Save as Image File

```{r save-chart, eval=FALSE}
# Create a chart
my_chart <- ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "skyblue") +
  labs(title = "Lemonade Sales") +
  theme_minimal()

# Save it!
ggsave("my_lemonade_chart.png", my_chart, width = 8, height = 6)

# It's now saved in your working directory!
```

---

# COMPREHENSIVE PROJECT: Pizza Shop Analysis

Let's analyze a week of pizza shop data!

```{r pizza-data}
# Create pizza shop data
pizza_shop <- tibble(
  day = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"),
  pizzas_sold = c(45, 52, 48, 55, 85, 120, 110),
  revenue = c(450, 520, 480, 550, 850, 1200, 1100),
  customers = c(30, 35, 32, 37, 55, 75, 70),
  day_type = c("Weekday", "Weekday", "Weekday", "Weekday", "Weekday", "Weekend", "Weekend")
)

pizza_shop
```

## Your Tasks

**Task 1:** Create a bar chart showing pizzas sold each day

```{r task-1}
# ANSWER:
ggplot(pizza_shop, aes(x = day, y = pizzas_sold)) +
  geom_col(fill = "tomato") +
  labs(
    title = "Pizzas Sold Each Day",
    x = "Day of Week",
    y = "Number of Pizzas Sold"
  ) +
  theme_minimal()

# Alternative with better ordering:
pizza_shop <- pizza_shop %>%
  mutate(day = factor(day, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

ggplot(pizza_shop, aes(x = day, y = pizzas_sold)) +
  geom_col(fill = "tomato") +
  labs(
    title = "Daily Pizza Sales",
    subtitle = "Weekend sales are significantly higher",
    x = "Day of Week",
    y = "Number of Pizzas Sold"
  ) +
  theme_light()
```

**Task 2:** Create a line chart showing revenue over the week

```{r task-2}
# ANSWER:
ggplot(pizza_shop, aes(x = day, y = revenue, group = 1)) +
  geom_line(color = "darkgreen", size = 1.5) +
  geom_point(color = "green", size = 4) +
  labs(
    title = "Weekly Revenue Trend",
    x = "Day of Week",
    y = "Revenue ($)"
  ) +
  theme_minimal()

# Alternative with enhanced styling:
ggplot(pizza_shop, aes(x = day, y = revenue, group = 1)) +
  geom_line(color = "darkgreen", size = 2) +
  geom_point(color = "gold", size = 5) +
  geom_area(alpha = 0.2, fill = "green") +
  labs(
    title = "Pizza Shop Revenue Over the Week",
    subtitle = "Saturday shows peak revenue",
    x = "Day of Week",
    y = "Revenue ($)"
  ) +
  theme_classic() +
  scale_y_continuous(labels = scales::dollar)
```

**Task 3:** Create a scatter plot: customers vs pizzas_sold

```{r task-3}
# ANSWER:
ggplot(pizza_shop, aes(x = customers, y = pizzas_sold)) +
  geom_point(size = 5, color = "navy") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Relationship: Customers vs Pizzas Sold",
    x = "Number of Customers",
    y = "Pizzas Sold"
  ) +
  theme_minimal()

# Alternative with labels:
library(ggrepel)
ggplot(pizza_shop, aes(x = customers, y = pizzas_sold)) +
  geom_point(size = 5, aes(color = day_type)) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  geom_text(aes(label = day), vjust = -1, size = 3) +
  labs(
    title = "Customer Count vs Pizzas Sold",
    subtitle = "Strong positive correlation",
    x = "Number of Customers",
    y = "Pizzas Sold",
    color = "Day Type"
  ) +
  theme_minimal()
```

**Task 4:** Make a bar chart colored by day_type (Weekday/Weekend)

```{r task-4}
# ANSWER:
ggplot(pizza_shop, aes(x = day, y = pizzas_sold, fill = day_type)) +
  geom_col() +
  labs(
    title = "Pizza Sales: Weekday vs Weekend",
    x = "Day of Week",
    y = "Pizzas Sold",
    fill = "Day Type"
  ) +
  theme_minimal()

# Alternative with custom colors:
ggplot(pizza_shop, aes(x = day, y = pizzas_sold, fill = day_type)) +
  geom_col() +
  labs(
    title = "Pizza Sales by Day Type",
    subtitle = "Weekend sales are more than double weekday sales",
    x = "Day of Week",
    y = "Pizzas Sold",
    fill = "Day Type"
  ) +
  scale_fill_manual(values = c("Weekday" = "lightblue", "Weekend" = "coral")) +
  theme_minimal() +
  theme(legend.position = "top")
```

**BONUS Challenge:** Combine all your charts into one display!

```{r task-bonus}
# ANSWER:

# Create all four charts
chart1 <- ggplot(pizza_shop, aes(x = day, y = pizzas_sold)) +
  geom_col(fill = "tomato") +
  labs(title = "Daily Sales", y = "Pizzas") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

chart2 <- ggplot(pizza_shop, aes(x = day, y = revenue, group = 1)) +
  geom_line(color = "darkgreen", size = 1.5) +
  geom_point(color = "green", size = 3) +
  labs(title = "Revenue Trend", y = "Revenue ($)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

chart3 <- ggplot(pizza_shop, aes(x = customers, y = pizzas_sold)) +
  geom_point(size = 4, color = "navy") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Customers vs Sales", x = "Customers", y = "Pizzas") +
  theme_minimal()

chart4 <- ggplot(pizza_shop, aes(x = day, y = pizzas_sold, fill = day_type)) +
  geom_col() +
  labs(title = "By Day Type", y = "Pizzas") +
  scale_fill_manual(values = c("Weekday" = "lightblue", "Weekend" = "coral")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Combine all four charts
(chart1 | chart2) / (chart3 | chart4)

# Alternative layout
# chart1 / (chart2 | chart3) / chart4
```

---

# Practice Problems

## Problem 1: Weather Data

```{r practice-weather}
weather <- tibble(
  month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun"),
  avg_temp = c(32, 35, 45, 55, 65, 75),
  rainfall = c(2.5, 2.0, 3.0, 3.5, 4.0, 3.0)
)

# ANSWERS:

# Task 1: Bar chart of temperature by month
ggplot(weather, aes(x = month, y = avg_temp)) +
  geom_col(fill = "orange") +
  labs(
    title = "Average Temperature by Month",
    x = "Month",
    y = "Temperature (°F)"
  ) +
  theme_minimal()

# Task 2: Line chart of rainfall
ggplot(weather, aes(x = month, y = rainfall, group = 1)) +
  geom_line(color = "blue", size = 1.5) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Rainfall",
    x = "Month",
    y = "Rainfall (inches)"
  ) +
  theme_minimal()

# Task 3: Scatter plot: temp vs rainfall
ggplot(weather, aes(x = avg_temp, y = rainfall)) +
  geom_point(size = 5, color = "purple") +
  labs(
    title = "Temperature vs Rainfall",
    x = "Average Temperature (°F)",
    y = "Rainfall (inches)"
  ) +
  theme_minimal()

# Bonus - all combined:
w1 <- ggplot(weather, aes(x = month, y = avg_temp)) +
  geom_col(fill = "orange") +
  labs(title = "Temperature") +
  theme_minimal()

w2 <- ggplot(weather, aes(x = month, y = rainfall, group = 1)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Rainfall") +
  theme_minimal()

w3 <- ggplot(weather, aes(x = avg_temp, y = rainfall)) +
  geom_point(size = 4, color = "purple") +
  labs(title = "Temp vs Rain") +
  theme_minimal()

(w1 | w2) / w3
```

## Problem 2: Student Grades

```{r practice-grades}
grades <- tibble(
  student = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  math = c(85, 92, 78, 95, 88),
  science = c(90, 85, 82, 98, 91)
)

# ANSWERS:

# Task 1: Bar chart of math grades
ggplot(grades, aes(x = student, y = math)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Math Grades by Student",
    x = "Student",
    y = "Grade"
  ) +
  theme_minimal()

# Task 2: Scatter plot: math vs science
ggplot(grades, aes(x = math, y = science)) +
  geom_point(size = 5, color = "darkred") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    title = "Math vs Science Grades",
    x = "Math Grade",
    y = "Science Grade"
  ) +
  theme_minimal()

# Task 3: Compare both subjects side by side
# First, reshape the data
grades_long <- grades %>%
  pivot_longer(
    cols = c(math, science),
    names_to = "subject",
    values_to = "grade"
  )

ggplot(grades_long, aes(x = student, y = grade, fill = subject)) +
  geom_col(position = "dodge") +
  labs(
    title = "Math and Science Grades Comparison",
    x = "Student",
    y = "Grade",
    fill = "Subject"
  ) +
  scale_fill_manual(values = c("math" = "lightblue", "science" = "lightgreen")) +
  theme_minimal()

# Alternative - faceted view:
ggplot(grades_long, aes(x = student, y = grade, fill = subject)) +
  geom_col() +
  facet_wrap(~subject) +
  labs(
    title = "Student Grades by Subject",
    x = "Student",
    y = "Grade"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# Quick Reference

## Basic Template

```r
ggplot(data = YOUR_DATA, aes(x = X_COLUMN, y = Y_COLUMN)) +
  geom_CHART_TYPE() +
  labs(title = "YOUR TITLE", x = "X LABEL", y = "Y LABEL")
```

## Common Chart Types

```r
# Bar Chart
geom_col()

# Line Chart
geom_line(group = 1)

# Scatter Plot
geom_point()

# Add trend line
geom_smooth(method = "lm")
```

## Common Customizations

```r
# Colors
geom_col(fill = "blue")
geom_point(color = "red")
geom_line(color = "green")

# Sizes
geom_line(size = 2)
geom_point(size = 5)

# Themes
+ theme_minimal()
+ theme_classic()
+ theme_bw()
+ theme_light()

# Labels
labs(title = "Title", x = "X axis", y = "Y axis")

# Combining charts
chart1 | chart2  # Side by side
chart1 / chart2  # Stacked
```

---

# Congratulations! 🎉

You can now create:
- ✅ Bar charts
- ✅ Line charts  
- ✅ Scatter plots
- ✅ Colored charts
- ✅ Combined charts
- ✅ Saved charts

**Keep practicing and have fun making beautiful visualizations!** 

---

# Session Information

```{r session-info}
sessionInfo()
```
