---
title: "Making Charts in R - Beginner Friendly"
author: "Student Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)
```

# Welcome to Data Visualization! ðŸ“Š

Learn to create beautiful charts and graphs in R. No experience needed!

## What You'll Learn

- Make basic charts (bar, line, scatter)
- Customize colors and labels
- Create multiple charts
- Save your charts

---

# Getting Started

```{r packages}
# Load packages
library(tidyverse)  # This includes ggplot2!

# Check if it loaded
cat("âœ“ Packages loaded successfully!\n")
```

---

# Create Sample Data

Let's make data about a lemonade stand!

```{r create-data}
# Weekly sales data
lemonade <- tibble(
  day = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"),
  cups_sold = c(25, 30, 35, 40, 50, 75, 80),
  temperature = c(70, 75, 80, 82, 85, 90, 88),
  price = c(2.00, 2.00, 2.50, 2.50, 2.50, 3.00, 3.00)
)

# Look at the data
lemonade
```

---

# PART 1: Bar Charts

Bar charts show comparisons between categories.

## Your First Chart! ðŸŽ‰

```{r first-chart}
# Basic bar chart
ggplot(data = lemonade, aes(x = day, y = cups_sold)) +
  geom_col()
```

**How this works:**
1. `ggplot()` - Start a chart
2. `data = lemonade` - Use this dataset
3. `aes(x = day, y = cups_sold)` - Put days on x-axis, sales on y-axis
4. `geom_col()` - Make it a bar chart

---

## Make It Better!

```{r better-bar}
ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "skyblue") +  # Color the bars
  labs(
    title = "Lemonade Sales by Day",
    x = "Day of Week",
    y = "Cups Sold"
  )
```

**ðŸŽ¯ YOUR TURN:** Change the color to "coral" or "lightgreen"

```{r practice-1-1}
# Your code here:

```

---

## Sort the Bars

```{r sorted-bar}
ggplot(lemonade, aes(x = reorder(day, cups_sold), y = cups_sold)) +
  geom_col(fill = "coral") +
  labs(title = "Sales by Day (Sorted)", x = "Day", y = "Cups Sold") +
  coord_flip()  # Make horizontal
```

**What's new?**
- `reorder(day, cups_sold)` - Sort by sales
- `coord_flip()` - Flip it sideways

**ðŸŽ¯ YOUR TURN:** Make a bar chart of price by day

```{r practice-1-2}
# Your code here:

```

---

# PART 2: Line Charts

Line charts show trends over time.

## Basic Line Chart

```{r line-basic}
ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_line(group = 1)  # group = 1 connects all points
```

**âš ï¸ Important:** For line charts with text on x-axis, add `group = 1`

---

## Make It Better!

```{r line-better}
ggplot(lemonade, aes(x = day, y = cups_sold, group = 1)) +
  geom_line(color = "blue", size = 1.5) +  # Thicker blue line
  geom_point(color = "red", size = 3) +    # Add red dots
  labs(
    title = "Lemonade Sales Trend",
    x = "Day",
    y = "Cups Sold"
  ) +
  theme_minimal()  # Clean look
```

**ðŸŽ¯ YOUR TURN:** Create a line chart showing temperature over the week

```{r practice-2-1}
# Your code here:

```

---

# PART 3: Scatter Plots

Scatter plots show relationships between two numbers.

## Basic Scatter Plot

```{r scatter-basic}
ggplot(lemonade, aes(x = temperature, y = cups_sold)) +
  geom_point()
```

**What this shows:** Do we sell more when it's hotter?

---

## Add a Trend Line

```{r scatter-trend}
ggplot(lemonade, aes(x = temperature, y = cups_sold)) +
  geom_point(size = 4, color = "darkblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add trend line
  labs(
    title = "Sales vs Temperature",
    x = "Temperature (Â°F)",
    y = "Cups Sold"
  )
```

**ðŸ’¡ What's `geom_smooth()`?**
- Adds a line showing the general trend
- `method = "lm"` means straight line
- `se = FALSE` removes the gray area

**ðŸŽ¯ YOUR TURN:** Make a scatter plot of price vs cups_sold

```{r practice-3-1}
# Your code here:

```

---

# PART 4: Colors and Themes

## Adding Colors by Category

```{r color-by-category}
# Add a weekend column
lemonade <- lemonade %>%
  mutate(
    is_weekend = if_else(day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
  )

# Color by weekend
ggplot(lemonade, aes(x = day, y = cups_sold, fill = is_weekend)) +
  geom_col() +
  labs(title = "Sales: Weekday vs Weekend", fill = "Day Type")
```

---

## Built-in Themes

```{r themes}
# Try different themes!
p <- ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "steelblue") +
  labs(title = "Lemonade Sales")

# Theme 1: Minimal (clean)
p + theme_minimal()

# Theme 2: Classic (traditional)
# p + theme_classic()

# Theme 3: Dark
# p + theme_dark()
```

**ðŸŽ¯ YOUR TURN:** Try `theme_bw()` or `theme_light()`

```{r practice-4-1}
# Your code here:

```

---

# PART 5: Multiple Charts Together

## Create Multiple Simple Charts

```{r multiple-charts}
library(patchwork)  # For combining charts

# Chart 1: Bar chart
p1 <- ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "coral") +
  labs(title = "Cups Sold") +
  theme_minimal()

# Chart 2: Line chart
p2 <- ggplot(lemonade, aes(x = day, y = temperature, group = 1)) +
  geom_line(color = "red", size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Temperature") +
  theme_minimal()

# Put them side by side
p1 | p2
```

**How it works:**
- Create each chart separately
- Use `|` to put them side by side
- Use `/` to put them on top of each other

---

## Stack Charts Vertically

```{r stack-charts}
# Put one above the other
p1 / p2
```

**ðŸŽ¯ YOUR TURN:** Create any two charts and combine them

```{r practice-5-1}
# Your code here:

```

---

# PART 6: Saving Your Charts

## Save as Image File

```{r save-chart, eval=FALSE}
# Create a chart
my_chart <- ggplot(lemonade, aes(x = day, y = cups_sold)) +
  geom_col(fill = "skyblue") +
  labs(title = "Lemonade Sales") +
  theme_minimal()

# Save it!
ggsave("my_lemonade_chart.png", my_chart, width = 8, height = 6)

# It's now saved in your working directory!
```

**ðŸ’¡ Tips:**
- Use `.png` for images
- Use `.pdf` for documents
- `width` and `height` are in inches

---

# COMPREHENSIVE PROJECT: Pizza Shop Analysis

Let's analyze a week of pizza shop data!

```{r pizza-data}
# Create pizza shop data
pizza_shop <- tibble(
  day = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"),
  pizzas_sold = c(45, 52, 48, 55, 85, 120, 110),
  revenue = c(450, 520, 480, 550, 850, 1200, 1100),
  customers = c(30, 35, 32, 37, 55, 75, 70),
  day_type = c("Weekday", "Weekday", "Weekday", "Weekday", "Weekday", "Weekend", "Weekend")
)

pizza_shop
```

## Your Tasks

**Task 1:** Create a bar chart showing pizzas sold each day

```{r task-1}
# Your code here:

```

**Task 2:** Create a line chart showing revenue over the week

```{r task-2}
# Your code here:

```

**Task 3:** Create a scatter plot: customers vs pizzas_sold

```{r task-3}
# Your code here:

```

**Task 4:** Make a bar chart colored by day_type (Weekday/Weekend)

```{r task-4}
# Your code here:

```

**BONUS Challenge:** Combine all your charts into one display!

```{r task-bonus}
# Your code here:

```

---

# Chart Type Guide

## When to Use Each Chart

**Bar Charts** ðŸ“Š
- Comparing categories
- Example: Sales by day, Products by type
- Use `geom_col()`

**Line Charts** ðŸ“ˆ
- Showing trends over time
- Example: Temperature, Stock prices
- Use `geom_line()`

**Scatter Plots** ðŸ”µ
- Showing relationships between two numbers
- Example: Height vs Weight, Price vs Sales
- Use `geom_point()`

---

# Quick Reference

## Basic Template

```r
ggplot(data = YOUR_DATA, aes(x = X_COLUMN, y = Y_COLUMN)) +
  geom_CHART_TYPE() +
  labs(title = "YOUR TITLE", x = "X LABEL", y = "Y LABEL")
```

## Common Customizations

```r
# Colors
geom_col(fill = "blue")
geom_point(color = "red")

# Sizes
geom_line(size = 2)
geom_point(size = 5)

# Themes
+ theme_minimal()
+ theme_classic()
+ theme_bw()

# Labels
labs(title = "Title", x = "X axis", y = "Y axis")
```

---

# Color Names You Can Use

Some easy colors to try:
- "red", "blue", "green"
- "pink", "purple", "orange"
- "lightblue", "darkgreen", "coral"
- "skyblue", "steelblue", "navy"
- "gold", "yellow", "lightyellow"

---

# Common Mistakes and Solutions

**Mistake 1:** Forgot the `+` sign
```r
# Wrong:
ggplot(data, aes(x = day, y = sales))
geom_col()

# Right:
ggplot(data, aes(x = day, y = sales)) +
  geom_col()
```

**Mistake 2:** Line chart not connecting
```r
# Add group = 1:
ggplot(data, aes(x = day, y = sales, group = 1)) +
  geom_line()
```

**Mistake 3:** Misspelled column names
- Check spelling with `colnames(your_data)`
- R is case-sensitive!

---

# Practice Problems

## Problem 1: Weather Data

```{r practice-weather}
weather <- tibble(
  month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun"),
  avg_temp = c(32, 35, 45, 55, 65, 75),
  rainfall = c(2.5, 2.0, 3.0, 3.5, 4.0, 3.0)
)

# Task 1: Bar chart of temperature by month
# Task 2: Line chart of rainfall
# Task 3: Scatter plot: temp vs rainfall

# Your code here:

```

## Problem 2: Student Grades

```{r practice-grades}
grades <- tibble(
  student = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  math = c(85, 92, 78, 95, 88),
  science = c(90, 85, 82, 98, 91)
)

# Task 1: Bar chart of math grades
# Task 2: Scatter plot: math vs science
# Task 3: Compare both subjects side by side

# Your code here:

```

---

# Congratulations! ðŸŽ‰

You can now create:
- âœ… Bar charts
- âœ… Line charts  
- âœ… Scatter plots
- âœ… Colored charts
- âœ… Combined charts
- âœ… Saved charts

**Keep practicing and have fun making beautiful visualizations!** 

---

# What's Next?

Want to learn more? Try:

1. **Advanced Chart Types**
   - Histograms (distribution of data)
   - Box plots (comparing groups)
   - Heatmaps (color grids)

2. **Interactive Charts**
   - Make charts you can click and hover
   - Use the `plotly` package

3. **Dashboards**
   - Put multiple charts on a web page
   - Share with others

---

# Helpful Resources

- [ggplot2 Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf)
- [R Graphics Cookbook](https://r-graphics.org/)
- [Data to Viz](https://www.data-to-viz.com/) - Which chart to use?

---

# Session Information

```{r session-info}
sessionInfo()
```
