---
title: "Dashboard Building Lab - ANSWER KEY"
author: "Answer Key"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Welcome to Dashboard Building! 📊

This document contains complete answers for the **Dashboard Building Lab**.

---

# Getting Started

## Load Packages

```{r packages}
library(flexdashboard)
library(tidyverse)
library(plotly)

cat("✓ Packages loaded!\n")
```

---

# Coffee Shop Dashboard - Complete Solution

Below is the complete solution for the Coffee Shop Dashboard exercise from the lab.

## The Data

```{r coffee-data}
coffee <- tibble(
  day = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"),
  drinks_sold = c(120, 140, 135, 150, 200, 250, 230),
  revenue = c(480, 560, 540, 600, 800, 1000, 920),
  customers = c(80, 95, 90, 100, 130, 160, 145)
)

coffee
```

---

## Complete Dashboard Code

**To use this code:**
1. Create a new R Markdown file in RStudio
2. Copy the entire code below (starting from the `---` YAML header)
3. Save it as `coffee_shop_dashboard.Rmd`
4. Click **"Run Document"** (NOT "Knit")
5. Your interactive dashboard will open in your browser!

### Full Dashboard Code:
`Lab_Beginner_Dashboard_Solution_ver1.Rmd`

---

# Alternative Dashboard Versions

## Version 2: Column Layout

An alternative version with a column-based layout: 
`Lab_Beginner_Dashboard_Solution_ver2.Rmd`

---

## Version 3: With Additional Features

This version includes more advanced features:
`Lab_Beginner_Dashboard_Solution_ver3.Rmd`

---

# Code Explanations

## Key Components Explained

### 1. YAML Header

```yaml
---
title: "Coffee Shop Dashboard ☕"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows    # or "columns"
    theme: flatly       # optional theme
runtime: shiny          # REQUIRED for interactivity
---
```

**Important:**
- `runtime: shiny` is REQUIRED for interactive dashboards
- `orientation` can be "rows" or "columns"
- Theme options: `default`, `cosmo`, `bootstrap`, `cerulean`, `journal`, `flatly`, `readable`, `spacelab`, `united`, `lumen`, `paper`, `sandstone`, `simplex`, `yeti`

---

### 2. Sidebar Creation

```r
Sidebar {.sidebar}
-----------------------------------------------------------------------

# Your sidebar content here
```

**Tips:**
- Use `.sidebar` class to create a sidebar
- Sidebars are great for controls and filters
- Can contain inputs, text, or any content

---

### 3. Input Controls

```r
# Dropdown menu
selectInput(
  "input_id",           # ID to reference later
  label = "Label:",     # Text shown to user
  choices = c("A", "B"), # Options
  selected = "A",       # Default selection
  multiple = TRUE       # Allow multiple selections
)

# Radio buttons
radioButtons(
  "input_id",
  label = "Choose one:",
  choices = c("Option 1" = "val1", "Option 2" = "val2"),
  selected = "val1"
)

# Slider
sliderInput(
  "input_id",
  label = "Select value:",
  min = 0,
  max = 100,
  value = 50,
  step = 5
)

# Checkbox
checkboxInput(
  "input_id",
  label = "Check this box",
  value = TRUE
)
```

---

### 4. Reactive Data

```r
# Create reactive data that updates when inputs change
filtered_data <- reactive({
  # Your filtering logic here
  if ("All" %in% input$day_select) {
    original_data
  } else {
    original_data %>%
      filter(day %in% input$day_select)
  }
})

# Use reactive data with () after the name
# CORRECT:
filtered_data()

# WRONG:
filtered_data
```

**Key Points:**
- Reactive expressions automatically update when inputs change
- Always use `()` when calling reactive data: `filtered_data()`
- Define reactive data once, use it multiple times

---

### 5. Value Boxes

```r
renderValueBox({
  # Calculate your value
  total <- sum(filtered_data()$column)
  
  # Create the value box
  valueBox(
    value = format(total, big.mark = ","),  # The number to display
    caption = "Total Sales",                # Description
    icon = "fa-dollar-sign",                # Font Awesome icon
    color = "success"                       # Color theme
  )
})
```

**Color Options:**
- `"primary"` - Blue
- `"success"` - Green
- `"info"` - Light Blue
- `"warning"` - Orange
- `"danger"` - Red

**Icon Options:** Use [Font Awesome](https://fontawesome.com/icons) icons with `fa-` prefix

---

### 6. Rendering Charts

```r
# Basic chart
renderPlotly({
  p <- ggplot(filtered_data(), aes(x = x_var, y = y_var)) +
    geom_col() +
    theme_minimal()
  
  ggplotly(p)
})

# With custom tooltip
renderPlotly({
  p <- ggplot(filtered_data(), aes(
    x = x_var, 
    y = y_var,
    text = paste("Day:", day, "<br>Value:", value)
  )) +
    geom_point(size = 3) +
    theme_minimal()
  
  ggplotly(p, tooltip = "text")
})
```

---

### 7. Rendering Tables

```r
# Simple table
renderTable({
  filtered_data() %>%
    select(Column1, Column2, Column3)
})

# Interactive table with DT package
renderDT({
  datatable(
    filtered_data(),
    options = list(
      pageLength = 10,
      searching = TRUE,
      ordering = TRUE
    )
  )
})
```

---

### 8. Layout Structure

```r
# Row-based layout
Row {data-height=150}
-----------------------------------------------------------------------

### Chart Title
# Chart code here

Row {data-height=400}
-----------------------------------------------------------------------

### Chart 1
# Code

### Chart 2
# Code


# Column-based layout
Column {data-width=350}
-----------------------------------------------------------------------

### Chart Title
# Chart code here

Column {data-width=650}
-----------------------------------------------------------------------

### Chart Title
# Chart code here
```

---

# Common Dashboard Patterns

## Pattern 1: Filter + Multiple Charts

```r
# 1. Create input
selectInput("filter", "Filter:", choices = unique(data$category))

# 2. Create reactive data
filtered <- reactive({
  data %>% filter(category == input$filter)
})

# 3. Use in multiple charts
renderPlotly({ 
  # Chart using filtered()
})

renderPlotly({ 
  # Another chart using filtered()
})
```

---

## Pattern 2: Multiple Filters Combined

```r
# Multiple inputs
selectInput("day", "Day:", choices = days)
selectInput("product", "Product:", choices = products)
sliderInput("price", "Min Price:", min = 0, max = 100, value = 0)

# Combine in reactive
filtered <- reactive({
  data %>%
    filter(
      day %in% input$day,
      product %in% input$product,
      price >= input$price
    )
})
```

---

## Pattern 3: Conditional Display

```r
# Checkbox to show/hide element
checkboxInput("show_trend", "Show trend line", value = TRUE)

# Use in chart
renderPlotly({
  p <- ggplot(data, aes(x, y)) + geom_point()
  
  if (input$show_trend) {
    p <- p + geom_smooth(method = "lm")
  }
  
  ggplotly(p)
})
```

---

# Troubleshooting Guide

## Common Issues and Solutions

### Issue 1: Dashboard Doesn't Update

**Problem:** Clicking filters doesn't update charts

**Solution:**
- Make sure `runtime: shiny` is in YAML header
- Check that you're using `filtered_data()` with `()` in render functions
- Verify reactive data is defined correctly

---

### Issue 2: Can't Find Function Error

**Problem:** `Error: could not find function "valueBox"`

**Solution:**
```r
# Make sure flexdashboard is loaded
library(flexdashboard)
```

---

### Issue 3: Charts Not Showing

**Problem:** Empty space where chart should be

**Solution:**
- Check for errors in R console
- Make sure data exists: `print(filtered_data())`
- Verify column names are correct
- Check that `group = 1` is included for line charts

---

### Issue 4: Layout Issues

**Problem:** Charts overlapping or wrong size

**Solution:**
```r
# Specify heights and widths
Row {data-height=200}
Column {data-width=350}

# Try different orientations
orientation: rows  # vs orientation: columns
```

---

# Best Practices

## Dashboard Design Tips

### 1. Keep It Simple
- Don't overcrowd with too many charts
- Use 3-6 main visualizations
- Leave white space

### 2. Use Hierarchy
```r
Row {data-height=150}  # Value boxes (small)
-----------------------------------------------------------------------

Row {data-height=400}  # Main charts (large)
-----------------------------------------------------------------------

Row {data-height=200}  # Tables (medium)
-----------------------------------------------------------------------
```

### 3. Color Consistency
```r
# Define color palette
colors <- c(
  "Weekday" = "#4682B4",
  "Weekend" = "#FF6347"
)

# Use throughout dashboard
scale_fill_manual(values = colors)
```

### 4. Meaningful Labels
```r
# Good
labs(
  title = "Weekly Sales Trend",
  x = "Day of Week",
  y = "Revenue ($)"
)

# Bad
labs(title = "Chart", x = "x", y = "y")
```

---

# Testing Checklist

Before sharing your dashboard:

- [ ] All filters work correctly
- [ ] Charts update when filters change
- [ ] Value boxes show correct calculations
- [ ] No error messages in console
- [ ] Tooltips work (hover over charts)
- [ ] Layout looks good on different screen sizes
- [ ] Table is readable and formatted
- [ ] Colors are distinguishable
- [ ] All labels are clear
- [ ] Data makes sense

---

# Publishing Options

## Option 1: Share RMD File
- Recipient needs R and RStudio
- Good for: Colleagues with R

## Option 2: shinyapps.io (Recommended)

```r
# Install if needed
install.packages("rsconnect")

# Connect account
library(rsconnect)
rsconnect::setAccountInfo(
  name = "your-account",
  token = "your-token",
  secret = "your-secret"
)

# Deploy
rsconnect::deployApp("path/to/dashboard.Rmd")
```

**Free Tier:**
- 5 applications
- 25 active hours/month
- Good for personal projects

## Option 3: RStudio Connect
- For organizations
- Scheduled updates
- User authentication

---

# Additional Resources

## Learning More

- [flexdashboard Documentation](https://pkgs.rstudio.com/flexdashboard/)
- [Shiny Tutorial](https://shiny.rstudio.com/tutorial/)
- [Shiny Gallery](https://shiny.rstudio.com/gallery/)
- [plotly R Documentation](https://plotly.com/r/)

## Helpful Functions

```r
# Debugging
print(filtered_data())      # Check reactive data
observeEvent(input$filter, {  # Watch input changes
  print(input$filter)
})

# Performance
# Cache expensive calculations
cached_data <- reactive({
  # Expensive operation
}) %>% bindCache(input$filter)
```

---

# Congratulations! 🎉

You now know how to:
- ✅ Create interactive dashboards
- ✅ Add filters and controls
- ✅ Use reactive data
- ✅ Display value boxes
- ✅ Combine multiple visualizations
- ✅ Deploy dashboards online

**Keep building amazing dashboards!** 📊✨

---

# Session Info

```{r session-info}
sessionInfo()
```
